---
title: "Variance by level — Will (Spanish formal) vs. Tan (Exp2, matched design)"
format: html
editor: visual
---

A simple descriptive of "noisiness": treat noise as the spread of responses, and look at that
spread at different **levels**. Tan here is Experiment 2 (AE-DLVOT), the matched multi-block
adaptation design, restricted to its test phase — the same 12 VOT steps, 3 minimal pairs, 3
exposure conditions, and 6 test blocks as Will, so the comparison is apples-to-apples. Two figures:
trial-level noise as a curve across VOT, and the between-level spreads (participants, items,
conditions, blocks) as single numbers.

Note: this first pass **skips the participant exclusions**, so Will shows ~129 participants rather
than the ~121 after exclusion. Treat the numbers as a first look.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

# --- same dataset reads as the same colour in every figure ---
colours.dataset <- c("Tan" = "#333333", "Will" = "#1E88E5")
```

# Check out variance

## Step 1 — Load both into one common frame

Each dataset reduces to the same seven columns (dataset, participant, VOT, item, condition, block,
response), so every level downstream is one grouped pipe. Tan's `ParticipantID` is coerced to
character so it stacks with Will's.

```{r load, message=FALSE, warning=FALSE}
# --- Tan = Exp2 (AE-DLVOT), test phase only ---
d.tan <- read_csv("../dataTan2025/data/TJ2025-experiment-results.csv", show_col_types = FALSE) |>
  filter(Phase == "test") |>
  transmute(
    dataset     = "Tan",
    participant = as.character(ParticipantID),
    VOT.ms      = as.numeric(Item.VOT),
    item        = Item.MinimalPair,
    condition   = Condition.Exposure,
    block       = as.character(Block),
    resp.t      = if_else(Response.Voicing == "voiceless", 1, 0)
  )

# --- Will = formal test blocks; map words to the same 3 minimal pairs ---
d.will <- read_csv("../data/formal/formal-full-data-preprocessed.csv", show_col_types = FALSE) |>
  filter(str_detect(trial_type, "test"), !is.na(resp.num)) |>
  transmute(
    dataset     = "Will",
    participant = as.character(ID),
    VOT.ms      = VOT,
    item        = case_when(
      word %in% c("dip", "tip")   ~ "diptip",
      word %in% c("din", "tin")   ~ "dintin",
      word %in% c("dill", "till") ~ "dilltill",
      TRUE                        ~ word
    ),
    condition   = participant.condition,
    block       = as.character(blockID),
    resp.t      = resp.num
  )

d.common <- bind_rows(d.tan, d.will) |> filter(!is.na(resp.t))

# sanity check: each should have 3 items, 3 conditions, 6 blocks, ~120 participants
d.common |>
  group_by(dataset) |>
  summarise(across(c(participant, item, condition, block), n_distinct), n.trials = n())
```

## Step 2 — Figure 1: trial noise as a curve

At each VOT step, one variance number — the spread of the 0/1 response there, which is p(1 - p).
Plotted along VOT, this traces the inverted-U: high near the boundary, ~0 at the clear ends.

```{r trial-noise}
d.trial.noise <- d.common |>
  group_by(dataset, VOT.ms) |>
  summarise(trial.variance = var(resp.t), .groups = "drop")
```

```{r plot-trial-noise, message=FALSE, warning=FALSE}
#| fig-cap: "Trial-level response noise. At each VOT step, the variance of the 0/1 'voiceless' response (= p(1-p)) for Tan (grey) and Will (blue). A clean dataset traces an inverted-U peaking near the category boundary and dropping to ~0 at the clear ends; a curve lifted off the floor at the extremes, or broader through the middle, indicates more lapsing / shallower categorization. Error bars omitted on this first pass."
plot.trial.noise <- d.trial.noise |>
  ggplot(aes(x = VOT.ms, y = trial.variance, colour = dataset)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = colours.dataset) +
  labs(x = "VOT (ms)", y = "Response variance at each VOT step") +
  theme_bw()

plot.trial.noise
```

## Step 3 — Between-level spreads

The building block for the three "between" levels is the same each time: summarize each unit by its
**mean proportion /t/**, then take the **variance across those unit means**. One small function,
called four times.

```{r between-levels}
# --- variance across the mean proportions of one grouping unit ---
between_level_variance <- function(data, unit_col) {
  data |>
    group_by(dataset, {{ unit_col }}) |>
    summarise(unit.mean = mean(resp.t), .groups = "drop") |>
    group_by(dataset) |>
    summarise(between.variance = var(unit.mean), .groups = "drop")
}

d.between <- bind_rows(
  between_level_variance(d.common, participant) |> mutate(level = "participants"),
  between_level_variance(d.common, item)        |> mutate(level = "items"),
  between_level_variance(d.common, condition)   |> mutate(level = "conditions"),
  between_level_variance(d.common, block)       |> mutate(level = "blocks")
) |>
  mutate(level = factor(level, levels = c("participants", "items", "conditions", "blocks")))

# sanity check: 4 levels x 2 datasets = 8 rows, all non-negative
d.between
```

## Step 4 — Figure 2: between-level bars

```{r plot-between-levels, message=FALSE, warning=FALSE}
#| fig-cap: "Between-level spread. For each level, the variance across that level's unit means (each unit summarized by its mean proportion 'voiceless'), for Tan (grey) and Will (blue). Higher = that level pulls responses apart more. Read the item, condition, and block bars gently — they are variances over only 3, 3, and 6 units respectively, so a bootstrap CI isn't legitimate; the participant bars (~120 units) are stable. The between-participant bar pools people across conditions, so it also absorbs a little of the condition effect."
plot.between.levels <- d.between |>
  ggplot(aes(x = level, y = between.variance, fill = dataset)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = colours.dataset) +
  labs(x = NULL, y = "Variance across unit means (proportion /t/)") +
  theme_bw()

plot.between.levels
```

# Check out variance at item level (minimal pair)

The item bar in Figure 2 stood out: Will's between-pair variance (~0.017) is ~15x Tan's (~0.001).
This section asks *why* — and separates a real perceptual/stimulus difference from an artifact of
imbalance, coding, or the skipped exclusions. Plan: (1) localize which pair(s) drive it, (2)
characterize *what kind* of difference it is (boundary shift vs slope vs endpoints), (3) rule out
three artifacts, (4) verdict.

## Item Step 1 — Localize: mean proportion /t/ per pair

If one pair is the outlier, the hunt narrows immediately.

```{r item-localize}
d.item.mean <- d.common |>
  group_by(dataset, item) |>
  summarise(mean.prop.t = mean(resp.t), n.trials = n(), .groups = "drop")

d.item.mean
```

## Item Step 2 — Characterize: per-pair curves, boundaries, slopes

Plot each pair's categorization curve, and put numbers on it with a crude per-pair boundary
(-intercept/slope from a logistic) and slope. Curves separated *horizontally* = a boundary
(calibration) shift; separated in *steepness* = one pair harder/noisier; separated at the *ends* =
a lapse difference.

```{r item-fit}
# --- one pair's logistic boundary (ms) and slope ---
fit_pair_boundary <- function(pair.trials) {
  model.pair <- glm(resp.t ~ VOT.ms, family = binomial, data = pair.trials)
  b <- coef(model.pair)
  tibble(boundary.ms = -b[["(Intercept)"]] / b[["VOT.ms"]], slope = b[["VOT.ms"]])
}

d.item.fit <- d.common |>
  group_by(dataset, item) |>
  group_modify(~ fit_pair_boundary(.x)) |>
  ungroup()

d.item.fit
```

```{r plot-item-curves, message=FALSE, warning=FALSE}
#| fig-cap: "Per-pair categorization curves. Proportion of 't'-responses vs VOT for each minimal pair (dill/till, din/tin, dip/tip), faceted by dataset; dashed line marks the 0.5 boundary. In Tan the three curves overlay tightly (boundaries within ~5 ms); in Will they fan out horizontally (boundaries ~35-56 ms), with din/tin shifted well to the left and dill/till to the right. Slopes are similar across Will's pairs, so the spread is a horizontal boundary shift, not a steepness difference. Error bars omitted."
colours.item <- c("dilltill" = "#E64A19", "dintin" = "#00897B", "diptip" = "#5E35B1")

d.item.curve <- d.common |>
  group_by(dataset, item, VOT.ms) |>
  summarise(prop.t = mean(resp.t), .groups = "drop")

plot.item.curves <- d.item.curve |>
  ggplot(aes(x = VOT.ms, y = prop.t, colour = item)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = colours.item) +
  facet_wrap(~ dataset) +
  labs(x = "VOT (ms)", y = "Proportion /t/", colour = "Minimal pair") +
  theme_bw()

plot.item.curves
```
```{r}
# --- give both datasets a common 1-6 test-block index (Tan blocks 1,3,5,7,8,9; Will T1..T6) ---
d.item.block <- d.common |>
  group_by(dataset) |>
  mutate(block.order = dense_rank(block)) |>
  ungroup() |>
  group_by(dataset, block.order, item, VOT.ms) |>
  summarise(prop.t = mean(resp.t), .groups = "drop")

# --- per-pair curves within each test block, dataset in rows, block in columns ---
colours.item <- c("dilltill" = "#E64A19", "dintin" = "#00897B", "diptip" = "#5E35B1")

plot.item.by.block <- d.item.block |>
  ggplot(aes(x = VOT.ms, y = prop.t, colour = item)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1) +
  scale_colour_manual(values = colours.item) +
  facet_grid(dataset ~ block.order) +
  labs(x = "VOT (ms)", y = "Proportion /t/", colour = "Minimal pair") +
  theme_bw()

plot.item.by.block
```
```{r}
# --- collapse to proportion /t/ per dataset x condition x block x pair x VOT ---
d.item.cond.block <- d.common |>
  group_by(dataset) |>
  mutate(block.order = dense_rank(block)) |>          # common 1-6 test-block index
  ungroup() |>
  group_by(dataset, condition, block.order, item, VOT.ms) |>
  summarise(prop.t = mean(resp.t), .groups = "drop")

colours.item <- c("dilltill" = "#E64A19", "dintin" = "#00897B", "diptip" = "#5E35B1")

# --- one condition x block facet grid for a single dataset ---
plot_item_grid <- function(ds, condition.levels) {
  d.item.cond.block |>
    filter(dataset == ds) |>
    mutate(condition = factor(condition, levels = condition.levels)) |>
    ggplot(aes(x = VOT.ms, y = prop.t, colour = item)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.7) +
    geom_point(size = 0.9) +
    scale_colour_manual(values = colours.item) +
    facet_grid(condition ~ block.order) +
    labs(x = "VOT (ms)", y = "Proportion /t/", colour = "Minimal pair",
         caption = paste0("Dataset: ", ds,
                          " — rows = exposure condition, columns = test block (1-6)")) +
    theme_bw()
}

plot.will.grid <- plot_item_grid("Will", c("expo-20ms", "expo-10ms", "expo+20ms"))
plot.tan.grid  <- plot_item_grid("Tan",  c("Shift0", "Shift10", "Shift40"))

plot.will.grid
plot.tan.grid
```

